Clustering
========================================================
### Cluster Western U.S. climate divisions by average winter precipitation.
Compute the average winter precipitation for each climate division in the Western U.S.
* The data will be a single column of one average precipitation value per division
```{r libs, results='hide', message=FALSE}
library(maps)
library(mapdata)
library(mclust)
library(ggplot2)
library(ncdf)
library(cluster)
```

```{r load-Precip}
## Load Climate Division Precipitation data (Western US)
px = scan("http://iridl.ldeo.columbia.edu/SOURCES/.NOAA/.NCDC/.CIRS/.ClimateDivision/lon/data.ch")
wi = px < -104
px = px[wi]
py = scan("http://iridl.ldeo.columbia.edu/SOURCES/.NOAA/.NCDC/.CIRS/.ClimateDivision/lat/data.ch")[wi]
cd.pcp = open.ncdf('pcp.cdf')
pi = get.var.ncdf(cd.pcp, 'IDIV')[wi]
ptm = get.var.ncdf(cd.pcp, 'T') / 12 + 1960
pcpm = get.var.ncdf(cd.pcp, 'pcp')[,wi]

## Compute winter average
yr<-trunc(ptm+1/24) # shift date up by two weeks so that Dec becomes part of next year; (e.g. yr runs Dec-Nov)
yr.list<-split(1:length(ptm), yr, drop=TRUE) # a list of which months go with which year.
djfm<-lapply(yr.list, '[', 1:4) # grab the first 4 months of the modified year 
wintavg<-sapply(djfm, function(i) colMeans(pcpm[i, ]))  #compute winter average (Dec-Mar)
rainw<-t(wintavg)
rainw<-rainw[-1,] # drop the first year
rainw<-rainw[-dim(rainw)[1],]
time<-as.numeric(row.names(rainw))
```

x is the data matrix to cluster
* columns are the attributes to cluster on (lat-long-precip)
* rows are records/observations to cluster (years)
```{r cluster-data}
## "rainw" contains winter precip for 81 climdivs in western U.S. for 1986-2013
x<-apply(X=rainw, MARGIN=2, FUN=mean) # 'climatology' for each climdiv
```
### Part (i) identify the number of clusters, K
* Select a desired number of clusters, say, j; 
* cluster the data and compute the WSS:
* repeat for j=1:10; 
* plot j versus WSS:
* select number of clusters K, where the WSS starts to saturate.

```{r wss-k-clusters, fig.align='left', fig.width=7, fig.height=4}
## K-means cluster
nk = 7  # number of clusters to consider
wss=1:nk
x<-as.data.frame(x)
wss <- (nrow(x)-1)*sum(apply(x,2,var))
for (i in 2:nk) wss[i] <- sum(kmeans(x,centers=i)$withinss)
plot(wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares") 
## --> select K corresponding to the WSS beyond which the reduction in variance is small. Similar to the Eigen spectrum (variogram).
```

```{r BIC-k-clusters, fig.height=9, fig.width=12}
## Alternatively, use BIC to find the best number of clusters.
d_clust <- Mclust(x, G=1:10)
m.best <- dim(d_clust$z)[2]
cat("model-based optimal number of clusters:", m.best, "\n")
par(mfrow=c(2,2), mar=c(5,4,2,2)+0.1)
plot(d_clust)
```
Methods above indicate that 3 groups is optimal.
* WSS levels off after 3 groups
* BIC plateaus after 3 groups
* Uncertainty is miniminzed at 3 groups
* Density is minimized at 3 groups

### Part (ii) cluster the data into K clusters and display them.
```{r cluster1-plot}
## Ward Hierarchical Clustering
d <- dist(x, method = "euclidean") # distance matrix
fit <- hclust(d, method="ward")
par(mfrow=c(1,1), mar=c(5,4,4,2)+0.1)
plot(fit) # display dendogram
groups <- cutree(fit, k=3) # cut tree into clusters
rect.hclust(fit, k=3, border="red") # draw dendogram with red borders around the clusters
```

### Part (iii) Repeat (i)-(ii) by including latitude and longitude – the data set will now be a matrix of 3 columns – precipitation, latitude and longitude.
```{r cluster2}
x<-data.frame(long=px, lat=py, rain=x)
```

```{r ref.label="wss-k-clusters", fig.width=7, fig.height=4, fig.align='left'}
```
Again, 3 groups is optimal.

Finally, we cluster the lat-long-precip data into K clusters and try a few visualizations
```{r cluster2-plots, message=FALSE}
## First, a generic spatial plot with respect to lat-long.
k <- 3
library(ggplot2)
clusters <- pam(x, k, stand=TRUE, cluster.only=T)
qplot(x=x$long, y=x$lat, color=factor(clusters)) + theme_bw() + scale_y_continuous(name="Latitude") + scale_x_continuous(name="Longitude")

## Next, try cluster plot against first 2 principal components
fit <- kmeans(x, 3)
library(cluster) 
clusplot(x, fit$cluster, color=TRUE, shade=TRUE, 
     labels=2, lines=0)

## Finally, try a centroid Plot against first 2 discriminant functions
library(fpc)
plotcluster(x, fit$cluster)
```